Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-07-29 pirouette_article.tex
Address that 'create_exemplary_dd_tree' creates trees with different numbers of tips
Processed all feedback. Thanks @rsetienne. Fix #113

# Create a DD phylogeny with 6 taxa and a crown age of 10
phylogeny <- create_exemplary_dd_tree()

@@

\begin{figure}[H]
  \includegraphics[width=0.9\textwidth]{pirouette_example_30/errors.png}
  \caption{
    \new{
      The impact of the tree prior for 
      the example tree in Figure \ref{fig:dd_tree}. 
      The alignment for this true tree was generated using a JC substitution model 
      and strict clock model. 
      For inferring the tree from this alignment in the 'generative' scenario
      the same substitution and clock models were used, 
      and a Yule tree prior (this is the assumed generative model,
      because the real generative model is of course unknown). 
      In the 'best' scenario the best-fitting candidate 
      models (TN substitution model, strict clock model and Yule tree prior) 
      were used. For the twin tree, the same models were chosen, 
      except that BD was used to generate the twin tree.
      The twin distributions show the baseline inference error.
    }
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_30}
\end{figure}

In the upper panel of Figure \ref{fig:example_30},
we can see that the error distributions of the \new{ (assumed) generative model (i.e. the known generative substitution and clock models, and the tree model that is assumed in inference of the true tree, and the tree model that is used for generating and inferring the twin tree)}
differ substantially between the true and twin tree. 
This difference shows the extent of the mismatch between
the true tree model (which is DD) and the (Yule) tree prior used in inference.
Because these distributions are distinctively different,
the inference error made when using an 
incorrect tree prior on a DD tree is quite profound.

Comparing the upper and lower panel of Figure \ref{fig:example_30}, 
we can see that the best
candidate model is only slightly better at inferring the true tree,
than the (assumed) generative model, 
thereby showing that they cannot compensate 
for the true generative tree model not being among the inference models.

\new{The candidate model that had highest evidence given the simulated alignment,
was TN, strict and BD (see Table \ref{tab:options} for the meaning of these 
abbreviations). 
The TN site model is a surprising result: it assumes nucleotide substitutions 
occur at different rates between translations and transversions 
and that nucleotides are available with different frequencies (that is,
a nucleotide may be rare or overabundant).
The strict clock model matches the model used to simulate the alignment.
The BD model is perhaps somewhat surprising for the true tre because the other alteranative standard tree prior, Yule, is probably closest to the true DD model because it shows no pull-of-the-present (but also no slowdown).
}

>>>>

# Create a DD phylogeny with 5 taxa and a crown age of 10
phylogeny <- create_exemplary_dd_tree()

@@

\begin{figure}[H]
  \includegraphics[width=0.9\textwidth]{pirouette_example_30/errors.png}
  \caption{
    \new{
      The impact of the tree prior for 
      the example tree in Figure \ref{fig:dd_tree}. 
      The alignment for this true tree was generated using a JC substitution model 
      and strict clock model. 
      For inferring the tree from this alignment in the 'generative' scenario,
      the same substitution and clock models were used, 
      and a Yule tree prior (this is the assumed generative model,
      because the real generative model is of course unknown).
      For the twin tree, the same inference model is used.
      In the 'best' scenarios, for the 'true' tree, 
      the best-fitting candidate models on the 'true' tree
      (JC substitution model, RLN clock model and BD tree prior) 
      was used. 
      In the 'best' scenarios, for the 'twin' tree, 
      the best-fitting candidate models on the 'twin' tree
      (JC substitution model, RLN clock model and Yule tree prior) 
      was used. 
      The twin distributions show the baseline inference error.
    }
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_30}
\end{figure}

In the upper panel of Figure \ref{fig:example_30},
we can see that the error distributions of the 
\new{ 
  (assumed) generative model (i.e. the known generative substitution and 
  clock models, and the tree model that is assumed in inference of the true 
  tree, and the tree model that is used for generating and inferring the 
  twin tree)
}
differ substantially between the true and twin tree. 
This difference shows the extent of the mismatch between
the true tree model (which is DD) and the (Yule) tree prior used in inference.
Because these distributions are distinctively different,
the inference error made when using an 
incorrect tree prior on a DD tree is quite profound.

Comparing the upper and lower panel of Figure \ref{fig:example_30}, 
we can see that 
\new{
  the best candidate model is slightly worse at inferring the true tree,
  than the (assumed) generative model, 
  indicating that the generative inference model we selected is a good pick.
}

\new{
  The candidate model that had highest evidence given the simulated alignment,
  was JC, RLN and BD (see Table \ref{tab:options} for the meaning of these 
  abbreviations). 
  The RLN clock model is a surprising result: it assumes nucleotide substitutions 
  occur at different rates between the taxa.
  The JC nucleotide substitution model matches the model used to simulate the alignment.
  The BD model is perhaps somewhat surprising for the true tree, 
  because the other alternative standard tree prior, 
  Yule, is probably closest to the true 
  DD model because it shows no pull-of-the-present (but also no slowdown).
}

### 2020-07-29 pirouette_supplement.tex
Processed all feedback. Thanks @rsetienne. Fix #113

Based on the 'true alignment' (center-left panel), the candidate model
with the highest marginal likelihood is determined, from a set of 
15 models. The set of models consists of all combinations
of all 4 nucleotides substitution models (JC, HKY, TN, GTR),
all 2 clock models (strict and relaxed log-normal) and 2 birth-death
models (Yule and Birth-Death), except the inference model used as the
generative model (JC, strict clock, Yule). The inference model that had the
highest evidence (as shown in Table \ref{tab:evidence_true}) was 
the inference model with a JC nucleotide substitution model,
a relaxed log-normal clock model and a Yule tree model.
The resulting posterior trees are shown in the second panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

Based on the 'twin alignment' (center-right panel), the candidate model
with the highest marginal likelihood is determined, from the same 
set of 15 candidate models.The inference model that had the
highest evidence (as shown in Table \ref{tab:evidence_twin}) was 
the inference model with a JC nucleotide substitution model,
a strict clock model and a BD tree model (note: this indeed matches how
the twin tree and twin tree were simulated).
From the 'twin alignment' this best candidate inference model is run.
The resulting posterior trees are shown in the fourth panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

The ESSes of the 'true' pipeline for the best candidate model
are shown in Table \ref{tab:esses_best}. 
From the names of the estimated parameters, it is clear that
the best candidate model has had a TN nucleotide substitution 
model (which can be inferred from the parameters'kappa1', 'kappa2', 'freqParameter.1', 
'freqParameter.2', 'freqParameter.3', 'freqParameter.4')
a strict clock (no parameter needed to be estimated) and a BD 
model ('BirthDeath', 'BDBirthRate' and 'BDDeathRate').

@@

From the names of the estimated parameters, it is clear that
the best candidate model for the twin tree
is JC nucleotide substitution 
model (no parameter needed to be estimated),
a strict clock (no parameter needed to be estimated) and a BD 
model ('BirthDate', 'BDBirthRate' and 'BDDeathRate').
Note that this matches the actual process of how the twin tree
and twin alignment are generated.

@@

The marginal likelihood (or evidence) data for the model comparison
performed in the 'true' pipeline is shown in Table \ref{tab:evidence_true}.
The best (that is, the one with the highest model weight)
candidate model assumes a TN nucleotide substitution model,
a strict clock and a BD model.

@@

The marginal likelihood (or evidence) data for model comparison
performed in the 'twin' pipeline is shown in Table \ref{tab:evidence_twin}.
The best (that is, the one with the highest model weight)
candidate model assumes a JC nucleotide substitution model,
a strict clock and a BD model.
Note that this matches the actual process of how the twin tree
and twin alignment are generated.

>>>>

Based on the 'true alignment' (center-left panel), the candidate model
with the highest marginal likelihood is determined, from a set of 
15 models. The set of models consists of all combinations
of all 4 nucleotides substitution models (JC, HKY, TN, GTR),
all 2 clock models (strict and relaxed log-normal) and 2 birth-death
models (Yule and Birth-Death), except the inference model used as the
generative model (JC, strict clock, Yule). 
The inference model that had the highest evidence 
(as shown in Table \ref{tab:evidence_true}) was 
the inference model with a JC nucleotide substitution model,
an RLN clock model and a BD tree model.
The resulting posterior trees are shown in the second panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

Based on the 'twin alignment' (center-right panel), the candidate model
with the highest marginal likelihood is determined, from the same 
set of 15 candidate models.The inference model that had the
highest evidence (as shown in Table \ref{tab:evidence_twin}) was 
the inference model with a JC nucleotide substitution model,
an RLN clock model and a Yule tree model (note: this indeed matches how
the twin tree and twin tree were simulated).
From the 'twin alignment' this best candidate inference model is run.
The resulting posterior trees are shown in the fourth panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

The ESSes of the 'true' pipeline for the best candidate model
are shown in Table \ref{tab:esses_best}. 
From the names of the estimated parameters, it is clear that
the best candidate model has had a JC nucleotide substitution 
model (no parameter needed to be estimated)
an RLN clock model (which can be inferred from the parameter 'rate.mean') 
and a BD model ('BirthDeath', 'BDBirthRate' and 'BDDeathRate').

@@

From the names of the estimated parameters, it is clear that
the best candidate model for the twin tree
is JC nucleotide substitution 
model (no parameter needed to be estimated),
an RLN clock model (which can be inferred from the parameter 'rate.mean') 
and a Yule model ('YuleModel', 'birthRate').
Note that there is a mismatch between the actual process of how the twin tree
and twin alignment are generated, as the twin tree is generated by a
BD process, and the alignment is simulated using a JC nucleotide
substitution model and a strict clock model.

@@

The marginal likelihood (or evidence) data for the model comparison
performed in the 'true' pipeline is shown in Table \ref{tab:evidence_true}.
The best (that is, the one with the highest model weight)
candidate model assumes a JC nucleotide substitution model,
an RLN clock and a BD tree model.

@@

The marginal likelihood (or evidence) data for model comparison
performed in the 'twin' pipeline is shown in Table \ref{tab:evidence_twin}.
The best (that is, the one with the highest model weight)
candidate model assumes a JC nucleotide substitution model,
an RLN clock and a Yule tree model.
Note that there is a mismatch between the actual process of how the twin tree
and twin alignment are generated, as the twin tree is generated by a
BD process, and the alignment is simulated using a JC nucleotide
substitution model and a strict clock model.
